---
title: "Project title"
subtitle: "Proposal"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   The dataset \`lemur_data.csv\` comes from [Kaggle.com](https://www.kaggle.com/datasets/jessemostipak/duke-lemur-center-data) and was released by Jesse Mostipak.

-   The dataset was orginally collected from the [2019 data release from the Duke Lemur Center Database](https://lemur.duke.edu/duke-lemur-center-database/) and further compiled by Zehr, SM, Roach RG, Haring D, Taylor J, Cameron FH, Yoder AD in their [research paper](https://www.nature.com/articles/sdata201419).

-   Animal data have been collected and entered by Duke Lemur Center staff according to standard operating procedures and USDA, AZA, and IACUC guidelines throughout the history of the center (United States Department of Agriculture, Association of Zoos and Aquariums, Institutional Animal Care and Use Committee respectively). Births, deaths, weights, enclosure moves, behaviors, and other significant events are recorded daily by animal care, veterinary, and research staff and subsequently entered into the permanent records by the DLC Registrar.

-   This dataset contains information on over 3,500 observations. Each observation represent a lemur, including lemur-information such as ancestry, reproduction, longevity, and body mass (in total 54 columns).

## Research question

-   What are the top 3 factors that influence the lifespan of lemurs?
-   Lemurs are the most endangered group of mammals. In fact, 98% of lemur species are endangered, and 31% of species are critically endangered nowadays. Therefore, it is crucial to conduct researches on the lifespan of lemurs to save the endangered species. Moreover, Duke Lemur Center has been the world leader in the study, care, and protection of lemurs since it was founded in 1966. As Duke students, we are able to utilize the resources at DLC to research on lemurs and potentially contribute to their studies.
-   Through our preliminary investigation, we found that the death age of lemurs varies a lot ranging from 0 to 35. Thus, we are interested in researching on what are the determining factors of lemurs' lifespan. Our hypotheses is that taxon, sex, weight are the top 3 factors that would affect the lifespan of lemurs.
-   We plan to research on all the variables within the dataset that could potentially affect lemur's lifespan. There are both categorical and quantitative variables involved in our research questions since categorical variables such as taxon and quantitative variables such as weight can all play a role in their lifespan. (include types of categorical and quantitative variables and examples)

## Glimpse of data

```{r}
#| label: load-lemur-data

lemur <- read_csv("data/lemur_data.csv")

glimpse(lemur)
```

# Data 2

## Introduction and data

-   The data comes from CORGIS (Collection of Really Great, Interesting, Situated Datasets) website.

-   The earthquake data was originally collected on 6/7/2016 by simply collecting information from the United States Geological Survey by Ryan Whitcomb.

-   Each observation contains a unique earthquake and all the information surrounding the circumstances of the earthquake, including but not limited to: coordinates of where the earthquake occurred, its magnitude, time of earthquake, and the depth of the earthquake.

-   Ethical concerns? (There probably aren't any)

## Research question

-   Are earthquakes stronger in places where they occur most often? Does the frequency of an earthquake's location have any correlation to its strength and magnitude? \-- It seems like the two questions kind of say the same thing. Also strength doesn't appear to be a variable. Could you use significance/depth/gap instead?
-   Also say why it is important (maybe for predicting the danger of earthquakes to save lives or something)
-   A rough look at the first 50 observations showed that a lot of earthquakes happened in Alaska and California. This made us wonder why this was the case and to see if those patterns are consistent throughout the rest of the data. Furthermore, because earthquakes happen more often in those states, we wonder if they are stronger than earthquakes that only occur once at a certain location. Our hypothesis is that location and magnitude have a positive correlation. --- you need to have 3 variables in the research question/hypothesis.
-   Our research question focuses on two things: location and magnitude. For location, we can consider the nominal categorical variable "location.name" which gives us the state/country in which the earthquake occurred. We can also consider "location.longitude" and "location.latitude" which gives us the exact coordinates of an earthquake's location, which would be continuous numerical variables. "Impact.magnitude" displays the magnitude of an earthquake which is a continuous numerical variable.

## Glimpse of data

```{r}
#| label: load-earthquake-data

earthquakes <- read.csv("data/earthquakes.csv")

glimpse(earthquakes)
```

# Data 3

## Introduction and data

-   The data is combined from Niche's "2023 Best Colleges in America" list

    (https://www.niche.com/colleges/search/best-colleges/) and the U.S. Department of Education.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data). The Niche data was scraped by Maia on October 17-19 2022.

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)
-   A description of the research topic along with a concise statement of your hypotheses on this topic.
-   Identify the types of variables in your research question. Categorical? Quantitative?

## Glimpse of data

```{r}
#| label: load-college-data

niche_data_500 <- read_csv("data/niche_data_500.csv")
# load DoE data

# colleges <- left_join(niche_data_500, ___)
```
