---
title: "Project title"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: true
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

## Introduction and data

-   Here are three potential research questions:

    1.  Which three variables have the greatest impact on a college's ranking?

-   These questions are important because students applying to college trust these rankings and weigh them into their college decisions. Due to large impact that a college has on a student's life, it is important to know where these ranking come from and what they actually measure.

-   In general, we want to examine how different variables affect a school's ranking on the Niche College Ranking List. We plan to look at variables that are typically thought to influence school rank such as average SAT score and acceptance rate, but we also want to look at variables that aren't typically thought of such as geographic region or endowment size.

    We hypothesize that: ACT/SAT test score, acceptance rate, and median earnings 10 years after graduation will have the biggest impact on college rank.

-   Here are the types of variables involved in our research questions:

    -   Categorical (nominal):

        -   geographic region

        -   accreditor institution

        -   control (public vs. private)

        -   Carnegie classification

    -   Numerical (continuous):

        -   admission rate

        -   Total share of enrollment by racial groups

        -   average net price

        -   average cost of attendance

        -   average faculty salary

        -   percentage of undergraduates who receive a Pell grant

        -   completion rate

        -   average age of entry

        -   share of female students

        -   share of married students

        -   share of first generation students

        -   median & mean family income

        -   median & mean earning 10 years after entry

        -   Value of school's endowment

    -   Numerical (discrete)

        -   mean (if there is one) and median ACT/SAT scores for sub-tests and composite

        -   Enrollment of undergraduate certificate/degree-seeking students

**Data Tidying**

```{r}
#| label: load-college-data

niche_data_500 <- read_csv("data/niche_data_500.csv")

glimpse(niche_data_500)

us_dep_of_ed <- read_csv("data/us_dep_of_ed.csv")

glimpse(us_dep_of_ed)
```

```{r}


us_dep_of_ed <- us_dep_of_ed |> 
  mutate(
    INSTNM = case_when(
      INSTNM == "Columbia University in the City of New York" ~ 
      "Columbia University",
      
      INSTNM == "Washington University in St Louis" ~ 
      "Washington University in St. Louis",
      
      INSTNM == "University of California-Los Angeles" ~ 
      "University of California - Los Angeles",
      
      INSTNM == "University of Michigan-Ann Arbor" ~ 
      "University of Michigan - Ann Arbor",
      
      INSTNM == "Georgia Institute of Technology-Main Campus" ~ 
      "Georgia Institute of Technology",
      
      INSTNM == "University of Virginia-Main Campus" ~ 
      "University of Virginia",
      
      INSTNM == "United States Military Academy" ~ 
      "United States Military Academy at West Point",
      
       INSTNM == "The University of Texas at Austin" ~ 
      "University of Texas - Austin",
      
       INSTNM == "University of California-Berkeley" ~ 
      "University of California - Berkeley",
      
      INSTNM == "University of California-Irvine" ~ 
      "University of California - Irvine",
      
      INSTNM == "University of California-San Diego" ~ 
      "University of California - San Diego",
      
      INSTNM == "University of California-Davis" ~ 
      "University of California - Davis",
      
      INSTNM == "University of California-Santa Barbara" ~ 
      "University of California - Santa Barbara",
      
      INSTNM == "University of California-Santa Cruz" ~ 
      "University of California - Santa Cruz",
      
      INSTNM == "University of California-Riverside" ~ 
      "University of California - Riverside",
      
      INSTNM == "University of North Carolina Wilmington" ~
      "University of North Carolina - Wilmington",
      
      INSTNM == "University of North Carolina at Greensboro" ~
      "University of North Carolina - Greensboro",
      
      INSTNM == "Albany College of Pharmacy and Health Sciences" ~
      "Albany College of Pharmacy & Health Sciences",
      
      INSTNM == "Arizona State University Campus Immersion" ~
      "Arizona State University",
      
      INSTNM == "Arizona State University-Downtown Phoenix" ~
      "Arizona State University - Downtown Phoenix Campus",
      
      INSTNM == "Augustana College" ~
      "Augustana College - Illinois",
      
      UNITID == "173160" ~
      "Bethel University - Minnesota",    
      
      INSTNM == "Binghamton University" ~
      "Binghamton University, SUNY", 
      
      INSTNM == "Bowling Green State University-Main Campus" ~
      "Bowling Green State University",    
      
      INSTNM == "California Polytechnic State University-San Luis Obispo" ~
      "California Polytechnic State University (Cal Poly) - San Luis Obispo", 
      
      INSTNM == "California State University-Fullerton" ~
      "California State University - Fullerton",    
      
      INSTNM == "California State University-Long Beach" ~
      "California State University - Long Beach", 
      
      INSTNM == "The College of Wooster" ~
      "College of Wooster",    
      
      INSTNM == "Colorado State University-Fort Collins" ~
      "Colorado State University", 
      
      INSTNM == "Concordia University-Wisconsin" ~
      "Concordia University - Wisconsin",    
      
      INSTNM == "CUNY Bernard M Baruch College" ~
      "CUNY Baruch College", 
      
      INSTNM == "CUNY City College" ~
      "CUNY City College of New York",    
      
      INSTNM == "D'Youville College" ~
      "D'Youville", 
      
      INSTNM == "Eastern New Mexico University-Main Campus" ~
      "Eastern New Mexico University",    
      
      INSTNM == "Embry-Riddle Aeronautical University-Daytona Beach" ~
      "Embry-Riddle Aeronautical University - Daytona Beach", 
      
      INSTNM == "Embry-Riddle Aeronautical University-Worldwide" ~
      "Embry-Riddle Aeronautical University - Worldwide", 
      
      INSTNM == "Florida Agricultural and Mechanical University" ~
      "Florida A&M University",    
      
      INSTNM == "Franklin and Marshall College" ~
      "Franklin & Marshall College",
      
      INSTNM == "Hobart William Smith Colleges" ~
      "Hobart and William Smith Colleges",
      
      INSTNM == "Indiana University-Bloomington" ~
      "Indiana University - Bloomington", 
      
      INSTNM == "Indiana University-East" ~
      "Indiana University - East",
      
      INSTNM == "Indiana University-Purdue University-Indianapolis" ~
      "Indiana University-Purdue University - Indianapolis (IUPUI)",
      
      INSTNM == "Indiana Wesleyan University-Marion" ~
      "Indiana Wesleyan University",
      
      INSTNM == "Keiser University-Ft Lauderdale" ~
      "Keiser University - Fort Lauderdale", 
      
      INSTNM == "Kent State University at Kent" ~
      "Kent State University",
      
      INSTNM == "Louisiana State University and Agricultural & Mechanical College" ~
      "Louisiana State University",
      
      UNITID == "151786" ~
      "Marian University Indianapolis",
      
      INSTNM == "Maryville University of Saint Louis" ~
      "Maryville University", 
      
      INSTNM == "Metropolitan State University" ~
      "Metropolitan State University - Minnesota",
      
      INSTNM == "Miami University-Oxford" ~
      "Miami University",
            
      INSTNM == "Minnesota State University-Mankato" ~
      "Minnesota State University, Mankato",
      
      INSTNM == "Molloy College" ~
      "Molloy University",
            
      INSTNM == "Monroe College" ~
      "Monroe College - Bronx/New Rochelle",
      
      INSTNM == "Mount Saint Mary's University" ~
      "Mount Saint Mary's University Los Angeles",
            
      INSTNM == "New Mexico State University-Main Campus" ~
      "New Mexico State University",
      
      INSTNM == "New Mexico Institute of Mining and Technology" ~
      "New Mexico Tech",
            
      INSTNM == "North Carolina State University at Raleigh" ~
      "North Carolina State University",
      
      INSTNM == "North Dakota State University-Main Campus" ~
      "North Dakota State University",
            
      UNITID ==  "154101" ~
      "Northwestern College - Iowa",
      
      INSTNM == "Ohio University-Main Campus" ~
      "Ohio University",
            
      INSTNM == "Oklahoma State University-Main Campus" ~
      "Oklahoma State University",
      
      INSTNM == "Pacific University" ~
      "Pacific University Oregon",
            
      INSTNM == "The Pennsylvania State University" ~
      "Penn State",
      
      INSTNM == "Pennsylvania State University-Penn State Fayette- Eberly" ~
      "Penn State Fayette, The Eberly Campus",
      
      INSTNM == "Pennsylvania State University-Penn State York" ~
      "Penn State York",
      
      INSTNM == "Purdue University-Main Campus" ~
      "Purdue University",
      
      INSTNM == "Rutgers University-New Brunswick" ~
      "Rutgers University - New Brunswick",
      
      INSTNM == "Rutgers University-Newark" ~
      "Rutgers University - Newark",
      
      INSTNM == "The University of the South" ~
      "Sewanee - The University of the South",
      
      INSTNM == "South Dakota School of Mines and Technology" ~
      "South Dakota School of Mines & Technology",
      
      INSTNM == "St Bonaventure University" ~
      "St. Bonaventure University",
      
      INSTNM == "Saint John Fisher College" ~
      "St. John Fisher University",
      
      INSTNM == "St. Joseph's University-New York" ~
      "St. Joseph's University - New York",
      
      INSTNM == "St Lawrence University" ~
      "St. Lawrence University",
      
      INSTNM == "St Olaf College" ~
      "St. Olaf College",
      
      INSTNM == "Stanbridge University" ~
      "Stanbridge University - Orange County",
      
      INSTNM == "Stephen F Austin State University" ~
      "Stephen F. Austin State University",
      
      INSTNM == "Stony Brook University" ~
      "Stony Brook University, SUNY",
      
      INSTNM == "SUNY College of Environmental Science and Forestry" ~
      "SUNY College of Environmental Science & Forestry",
      
      INSTNM == "Farmingdale State College" ~
      "SUNY Farmingdale State College",
      
      INSTNM == "State University of New York at New Paltz" ~
      "SUNY New Paltz",
      
      INSTNM == "Texas A & M University-College Station" ~
      "Texas A&M University",
      
      INSTNM == "American Musical and Dramatic Academy" ~
      "The American Musical and Dramatic Academy (AMDA) - New York",
      
      INSTNM == "The College of Saint Scholastica" ~
      "The College of St. Scholastica",
      
      INSTNM == "Cooper Union for the Advancement of Science and Art" ~
      "The Cooper Union for the Advancement of Science and Art",
      
      INSTNM == "The Master's University and Seminary" ~
      "The Master's University",
      
      INSTNM == "Ohio State University-Main Campus" ~
      "The Ohio State University",
      
      INSTNM == "University of Alabama in Huntsville" ~
      "The University of Alabama in Huntsville",
      
      INSTNM == "University of Baltimore" ~
      "The University of Baltimore",
      
      INSTNM == "University of Tulsa" ~
      "The University of Tulsa",
      
      INSTNM == "University of Virginia's College at Wise" ~
      "The University of Virginia's College at Wise",
      
      INSTNM == "Trinity College" ~
      "Trinity College - Connecticut",
      
      INSTNM == "Tulane University of Louisiana" ~
      "Tulane University",
      
      UNITID == "196866" ~
      "Union College - New York",
      
      INSTNM == "University at Buffalo" ~
      "University at Buffalo, SUNY",
      
      INSTNM == "University of Alabama at Birmingham" ~
      "University of Alabama - Birmingham",
      
      INSTNM == "University of Cincinnati-Main Campus" ~
      "University of Cincinnati",
      
      INSTNM == "University of Colorado Denver/Anschutz Medical Campus" ~
      "University of Colorado Denver",
      
      INSTNM == "University of Maryland-College Park" ~
      "University of Maryland - College Park",
      
      INSTNM == "University of Maryland-Baltimore County" ~
      "University of Maryland, Baltimore County",
      
      INSTNM == "University of Massachusetts-Amherst" ~
      "University of Massachusetts - Amherst",
      
      INSTNM == "University of Massachusetts-Lowell" ~
      "University of Massachusetts Lowell",
      
      INSTNM == "University of Michigan-Dearborn" ~
      "University of Michigan - Dearborn",
      
      INSTNM == "University of Minnesota-Crookston" ~
      "University of Minnesota Crookston",
      
      INSTNM == "University of Minnesota-Duluth" ~
      "University of Minnesota Duluth",
      
      INSTNM == "University of Minnesota-Twin Cities" ~
      "University of Minnesota Twin Cities",
      
      INSTNM == "University of Missouri-Columbia" ~
      "University of Missouri",
    
      INSTNM == "Embry-Riddle Aeronautical University-Prescott" ~
      "Embry-Riddle Aeronautical University - Prescott",
      
      INSTNM == "South Dakota State University" ~
      "South Dakota State University",
      
      INSTNM == "St Catherine University" ~
      "St. Catherine University",
      
      INSTNM == "University of Missouri-Kansas City" ~
      "University of Missouri - Kansas City", 
      
      INSTNM == "University of Missouri-St Louis" ~
      "University of Missouri - St. Louis", 
      
      INSTNM == "University of Nebraska-Lincoln" ~
      "University of Nebraska - Lincoln", 
      
      INSTNM == "University of Nevada-Reno" ~
      "University of Nevada - Reno", 
      
      INSTNM == "University of New Hampshire-Main Campus" ~
      "University of New Hampshire",
      
      INSTNM == "University of Oklahoma-Norman Campus" ~
      "University of Oklahoma",
      
      INSTNM == "University of Pittsburgh-Pittsburgh Campus" ~
      "University of Pittsburgh", 
      
      INSTNM == "University of South Carolina-Columbia" ~
      "University of South Carolina", 
      
      INSTNM == "University of St Francis" ~
      "University of St. Francis - Illinois",
      
      UNITID == "174914" ~
      "University of St. Thomas - Minnesota",
      
      UNITID == "227863" ~
      "University of St. Thomas - Texas",    
      
      INSTNM == "The University of Tennessee-Knoxville" ~
      "University of Tennessee", 
      
      INSTNM == "The University of Tennessee-Martin" ~
      "University of Tennessee at Martin", 
      
      INSTNM == "The University of Texas at Arlington" ~
      "University of Texas - Arlington", 
      
      INSTNM == "The University of Texas at Dallas" ~
      "University of Texas - Dallas",
      
      INSTNM == "The University of Texas Permian Basin" ~
      "University of Texas - Permian Basin",
      
      INSTNM == "The University of Texas Rio Grande Valley" ~
      "University of Texas - Rio Grande Valley", 
      
      INSTNM == "The University of Texas at Tyler" ~
      "University of Texas - Tyler", 
      
      INSTNM == "University of Washington-Seattle Campus" ~
      "University of Washington", 
      
      INSTNM == "University of Washington-Bothell Campus" ~
      "University of Washington - Bothell", 
      
      INSTNM == "University of Washington-Tacoma Campus" ~
      "University of Washington - Tacoma",
      
      INSTNM == "The University of West Florida" ~
      "University of West Florida",
      
      INSTNM == "University of Wisconsin-Madison" ~
      "University of Wisconsin", 
      
      INSTNM == "University of Wisconsin-La Crosse" ~
      "University of Wisconsin - La Crosse", 
      
      INSTNM == "Virginia Polytechnic Institute and State University" ~
      "Virginia Tech", 
      
      INSTNM == "West Texas A & M University" ~
      "West Texas A&M University", 
      
      UNITID == "230807" ~
      "Westminster College - Utah",
      
      INSTNM == "Wheaton College" ~
      "Wheaton College - Illinois",
      
      INSTNM == "Wheaton College (Massachusetts)" ~
      "Wheaton College - Massachusetts", 
      
      TRUE ~ INSTNM
      )
  )

colleges <- niche_data_500 |> 
  left_join(us_dep_of_ed, by = c("college" ="INSTNM")) |> 
  filter(rank != 141) |> 
  filter(rank != 181)

write_csv(colleges, "data/colleges.csv")

```

The Observations for `University of South Florida - Sarasota-Manatee` and `University of South Florida - St. Petersburg` have been dropped from the `colleges` data set due to their non-existence in the `us_dep_of_ed` data set. They were not present in the data set because these two universities were combined with the main University of South Florida campus. This is why the `colleges` data set only has 498 observations

## Methodology

The methodology section should include visualizations and summary statistics relevant to your research question. You should also justify the choice of statistical method(s) used to answer your research question.

### Prepare Data

1.  Pick variables that make sense to focus on

2.  Standardize all the continuous variables to z-scores, mean = 0, standard devation = 1

3.  Check correlation coefficients between all the the all the exponential variables. For the ones that have a r \> 0.8, pick one to put in the model. We can create multiple models with different options.

### Build model (Iterations)

1.  Put all of our chosen variables into a model

2.  Remove variables with the lowest slope coefficients until we get the best adjusted r-squared for the overall models.

3.  Look at slope coefficients for variables in models for variables in the model. The one with the highest slope is the most significant.

## Results

Showcase how you arrived at answers to your research question using the techniques we have learned in class (and beyond, if you're feeling adventurous).

Provide only the main results from your analysis. The goal is not to do an exhaustive data analysis (calculate every possible statistic and perform every possible procedure for all variables). Rather, you should demonstrate that you are proficient at asking meaningful questions and answering them using data, that you are skilled in interpreting and presenting results, and that you can accomplish these tasks using R. More is not better.
