---
title: "College Ranking Predictions"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: false
editor_options: 
  chunk_output_type: inline
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
library(corrplot)
library(tidymodels)
```

## Introduction and data

### Research Background

Every year, millions of students apply to colleges across the United States, and many of them use college rankings lists from sources such as *US News and World Report*, Forbes.com, and Niche.com to to help them decide where to apply and where to go. In recent years, these lists have been heavily criticized for focusing on "exclusivity and resources, rather than accessibility and economic mobility" (1).The system can also be easily manipulated by a university if that university prioritize certain metrics to raise their rankings, as seen in Northeastern University meteoric rise from #163 to #49 on the US News and World Report list in only 17 years. The president of Northeastern University even explicitly stated that it a top priority of the university to raise its ranking (2).

\(1\) https://thehill.com/changing-america/enrichment/education/3641004-the-scandal-facing-college-ranking-lists-explained/

\(2\) https://www.bostonmagazine.com/news/2014/08/26/how-northeastern-gamed-the-college-rankings/

These lists are important because students applying to college trust these rankings and weigh them into their college decisions. Due to the large impact that a college has on a student's life, it is important to know where these rankings come from and what they actually measure. In this project, we will explore how influential different metrics are in determining a college's ranking.

In general, we want to examine how different variables affect a school's ranking on the **Niche College Ranking List**. We plan to look at variables that are typically thought to influence school rank such as average SAT score and admission rate, but we also want to look at variables that aren't typically thought of such as geographic region or endowment size.

### Research Questions and Hypothesis

**Question:** What variables (and what types of variables) have the greatest influence on a school's place in the Niche College Ranking List?

**Hypothesis:** We hypothesize that: ACT/SAT test score, acceptance rate, and median earnings 10 years after graduation will have the biggest impact on college rank.

In our project, we will join two data sets: The Niche College Rankings list and the US Department of Education College Scorecard

### Data

#### Data Set #1: Niche

-   The first data set comes from Niche's "2023 Best Colleges in America" list

    (https://www.niche.com/colleges/search/best-colleges/)

-   Niche aggregates data from a variety of sources, including the US Department of Education and reviews from students and alumni, to build their list of college rankings. The rankings list is updated monthly to reflect new data that Niche receives; although, data from the US Department o Education is only received on an annual basis. The Niche data was scraped by Maia on October 17-19 2022.

-   There are 500 observations, representing the top 500 schools in the United States. Each observation has two variables: `college` and `rank`.

#### Data Set #2: US Department of Education

-   The second data set comes from the US Department of Education's College Scorecard, which is an exhaustive summary of characteristics and statistics for all colleges and universities in the United States.

    (https://collegescorecard.ed.gov/data/)

-   The College Scorecard is updated by the Education Department as it collects new data. Data used in the scorecard comes comes from data reported by the institutions, data on federal financial aid, data from taxes, and data from other federal agencies.

-   There were 2,989 variables in the original data set, many of which we don't need to answer our question, and since this data set was too large to load into RStudio, we used Excel to narrow this to 63 variables.There are 6681 observations in the data set, representing all of the colleges and universities in the United States.

### Data summary

Here is a summary of the data we will be using in or analysis.

+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| Variable Name | Categorical (C)  | Description                                                                                                        | Levels of Cat. Variable                                                                                     |
|               |                  |                                                                                                                    |                                                                                                             |
|               | Quantitative (Q) |                                                                                                                    |                                                                                                             |
+===============+==================+====================================================================================================================+=============================================================================================================+
| college       | C                | Institution name                                                                                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| rank          | Q                | Niche rank                                                                                                         |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| REGION        | C                |                                                                                                                    | New England, Mid East, Great Lakes, Plains, Southeast, Southwest, Rocky Mountains, Far West, Outlying Areas |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACCREDAGENCY  | C                | Accreditor for Institution                                                                                         |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| CONTROL       | C                |                                                                                                                    | Public, Private nonprofit, private for-profit                                                               |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| CCBASIC       | C                | Carnegie Classification \--basic                                                                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ADM_RATE      | Q                | Admission rate                                                                                                     |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS          | Q                | Enrollment of undergraduate certificate/degree-seeking students                                                    |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_WHITE    | Q                | Total share of enrollment of undergraduate degree-seeking students who are white                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_BLACK    | Q                | Total share of enrollment of undergraduate degree-seeking students who are black                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_HISP     | Q                | Total share of enrollment of undergraduate degree-seeking students who are Hispanic                                |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_ASIAN    | Q                | Total share of enrollment of undergraduate degree-seeking students who are Asian                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_AIAN     | Q                | Total share of enrollment of undergraduate degree-seeking students who are American Indian/Alaska Native           |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_NHPI     | Q                | Total share of enrollment of undergraduate degree-seeking students who are Native Hawaiian/Pacific Islander        |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_2MOR     | Q                | Total share of enrollment of undergraduate degree-seeking students who are two or more races                       |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_NRA      | Q                | Total share of enrollment of undergraduate degree-seeking students who are non-resident aliens                     |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_UNKN     | Q                | Total share of enrollment of undergraduate degree-seeking students whose race is unknown                           |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| NPT4_PUB      | Q                | Average net price for Title IV institutions (public institutions)                                                  |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| NPT4_PRIV     | Q                | Average net price for Title IV institutions (private for-profit and nonprofit institutions)                        |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| COSTT4_A      | Q                | Average cost of attendance (academic year institutions)                                                            |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| COSTT4_P      | Q                | Average cost of attendance (program-year institutions)                                                             |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| AVGFACSAL     | Q                | Average faculty salary                                                                                             |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| PCTPELL       | Q                | Percentage of undergraduates who receive a Pell Grant                                                              |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| C150_4        | Q                | Completion rate for first-time, full-time students at four-year institutions (150% of expected time to completion) |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| AGE_ENTRY     | Q                | Average age of entry                                                                                               |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| FEMALE        | Q                | Share of female students                                                                                           |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| MARRIED       | Q                | Share of married students                                                                                          |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| FIRST_GEN     | Q                | Share of first-generation students                                                                                 |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| FAMINC        | Q                | Average family income                                                                                              |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| MD_FAMINC     | Q                | Median family income                                                                                               |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ENDOWBEGIN    | Q                | Value of school's endowment at the beginning of the fiscal year                                                    |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| SAT_AVG       | Q                | Average SAT equivalent score of students admitted                                                                  |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| SATVRMID      | Q                | Midpoint of SAT scores at the institution (critical reading)                                                       |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| SATMTMID      | Q                | Midpoint of SAT scores at the institution (math)                                                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| SATWRMID      | Q                | Midpoint of SAT scores at the institution (writing)                                                                |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACTCMMID      | Q                | Midpoint of the ACT cumulative score                                                                               |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACTENMID      | Q                | Midpoint of the ACT english score                                                                                  |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACTMTMID      | Q                | Midpoint of the ACT math score                                                                                     |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACTWRMID      | Q                | Midpoint of the ACT writing score                                                                                  |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+

## Methodology

### Scrape Data

Scrape the data Niche.com and download data from the US Department of Education, and import.

The steps were done in an R script titled `niche-scrape.R`

### Import Dataset

```{r}
#| label: import
niche_data_500 <- read_csv("data/niche_data_500.csv")
us_dep_of_ed <- read_csv("data/us_dep_of_ed.csv", na = c("NULL", "PrivacySuppressed"))
```

### Join Data

1.  

2.  Change college names in Department of Education file to match Niche file

3.  Left join the `us_dep_of_ed` data set to Niche data set (`colleges`) by school name to make a new data set with 500 observations and 64 variables.

4.  The Observations for `University of South Florida - Sarasota-Manatee` and `University of South Florida - St. Petersburg` have been dropped from the `colleges` data set due to their non-existence in the `us_dep_of_ed` data set. They were not present in the data set because these two universities were combined with the main University of South Florida campus. This is why the `colleges` data set only has 498 observations

## Tidy Data

These steps were done in an R script titled `create-datasets.R`

1.  Pick variables that make sense to focus on

```{=html}
<!-- -->
```
2.  Standardize all the continuous variables to z-scores, mean = 0, standard deviation = 1

```{r}
#| label: load-data

colleges <- read_csv("data/colleges.csv")
scaled_data <- read_csv("data/scaled_data.csv")
```

**Summarize and Visualize**

```{r}

colleges_levels <- colleges |> 
  mutate(
    level = case_when(
      rank <= 100 ~ "1 to 100",
      rank > 100 & rank <= 200 ~ "101 to 200",
      rank > 200 & rank <= 300 ~ "201 to 300",
      rank > 300 & rank <= 400 ~ "301 to 400",
      rank > 400 & rank <= 498 ~ "401 to 498",
    )
  )

colleges_levels |> 
  group_by(level) |> 
  summarize(
    mean_ADM_Rate = mean(ADM_RATE, na.rm = T),
    mean_SAT_AVG = mean(SAT_AVG, na.rm = T),
    mean_ACTCMMID = mean(ACTCMMID, na.rm = T),
    mean_UGDS_WHITE = mean(UGDS_WHITE, na.rm = T),
    mean_UGDS_ASIAN = mean(UGDS_ASIAN, na.rm = T),
    mean_COSTT4_A = mean(COSTT4_A, na.rm = T)
  )




```

```{r}

#| label: rank-vs-ADM_RATE

colleges |> 
  ggplot(aes(x = ADM_RATE, y = rank)) +
  geom_point()

```

```{r}

#| label: ADM_RATE-vs-rank-stand

scaled_data |> 
  ggplot(aes(x = ADM_RATE, y = rank)) +
  geom_point()
```

```{r}

#| label: rank-vsADM_RATE-by-CCBASIC

scaled_data |> 
  ggplot(aes(x = ADM_RATE, y = rank, color = CCBASIC)) +
  geom_point()

```

```{r}
#| label: ADM_RATE-vs-rank-by-region

scaled_data |> 
  ggplot(aes(x = ADM_RATE, y = rank, color = CONTROL)) +
  geom_point()
```

```{r}

#| label: ADM_RATE-vs-rank-by-ACCREDAGENCY

scaled_data |> 
  ggplot(aes(x = ADM_RATE, y = rank, color = ACCREDAGENCY)) +
  geom_point(show.legend = F)

```

```{r}

colleges |> 
  ggplot(aes(x = UGDS_WHITE, y = rank)) +
  geom_point()

colleges |> 
  ggplot(aes(x = UGDS_ASIAN, y = rank)) +
  geom_point()

colleges |> 
  ggplot(aes(x = UGDS_ASIAN, y = log(rank))) +
  geom_point()

colleges |> 
  ggplot(aes(x = UGDS_BLACK, y = rank)) +
  geom_point()

colleges |> 
  ggplot(aes(x = UGDS_BLACK, y = log(rank))) +
  geom_point()

```

```{r}

scaled_data |> 
  ggplot(aes(x = FIRST_GEN, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = UGDS, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = NPT4_PRIV, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = NPT4_PUB, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = NPT4_PUB, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = COSTT4_A, y = rank, color = CONTROL)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = AVGFACSAL, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = PCTPELL, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = C150_4, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = AGE_ENTRY, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = FEMALE, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = MARRIED, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = FIRST_GEN, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = FAMINC, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = MD_FAMINC, y = rank)) +
  geom_point()

scaled_data |> 
  ggplot(aes(x = ENDOWBEGIN, y = rank)) +
  geom_point()


```

```{r}

#| label: rank-vs-SAT_AVG

colleges |> 
  ggplot(aes(x = SAT_AVG, y = rank)) +
  geom_point()
```

```{r}

#| label: rank-vs-ACTCMMID

colleges |> 
  ggplot(aes(x = ACTCMMID, y = rank)) +
  geom_point()

```

3.  Check correlation coefficients between all the all the exponential variables. For the ones that have a r \> 0.8, pick one to put in the model. We can create multiple models with different options.

    We remove the columns that has more than 100 NA and leave the rest for calculating correlations.

    ```{r}
    #| label: correlations

    data_for_corr <- scaled_data |>
      select(is.numeric)


    correlations = data_for_corr |>
      select(-c("NPT4_PUB", "NPT4_PRIV", "COSTT4_P", "SATWRMID", "SATVRMID", "SATMTMID", "ACTENMID", "ACTMTMID", "ACTWRMID")) |>
        na.omit() |>
        cor() 

    correlations |>
        corrplot(method="color")


    # correlations <- dataset |>
    #   na.omit() |>
    #   cor()
      

    correlations[correlations < 0.8 | correlations == 1] <- ""
    correlations
    ```

According to the correlation matrix, we need to remove variable pairs that have a high correlation (the absolute value \> 0.8).

The second output sections only shows the correlation entries with a value greater than 0.8. These variable pairs are (MD_FAMINC, FAMINC), (C150_4, SAT_AVG), (C150_4, ACTCMMID) and (ACTCMMID, SAT_AVG). Therefore, we will drop the variables C150_4, MD_FAMINC, ACTCMMID and preserve SAT_AVG and FAMINC to represent all other variables.

### Build model (Iterations)

1.  Put all of our chosen variables into a model

```{r}

data_set <- scaled_data |>
  select(c("ADM_RATE", "UGDS", "UGDS_WHITE", "UGDS_BLACK", "UGDS_HISP", "UGDS_ASIAN", "UGDS_AIAN", "UGDS_NHPI", "UGDS_2MOR", "UGDS_NRA", "UGDS_UNKN", "COSTT4_A", "AVGFACSAL", "PCTPELL", "AGE_ENTRY", "FAMINC", "ENDOWBEGIN", "SAT_AVG", "rank", "REGION", "CONTROL", "CCBASIC", "college", "ACCREDAGENCY", "FEMALE", "MARRIED", "FIRST_GEN"))

data_set
```

```{r}
# data_set_fit <- linear_reg() |>
#   fit(rank ~ ADM_RATE + UGDS + UGDS_WHITE + UGDS_BLACK + UGDS_HISP + UGDS_ASIAN + UGDS_AIAN + UGDS_NHPI + UGDS_2MOR + UGDS_NRA + UGDS_UNKN+ COSTT4_A + AVGFACSAL + PCTPELL + AGE_ENTRY + FAMINC + ENDOWBEGIN + SAT_AVG + rank + REGION + CONTROL + CCBASIC + college + ACCREDAGENCY, data = data_set)
# 
# tidy(data_set_fit)

data_set_fit <- linear_reg() |>
  fit(rank ~ ADM_RATE + UGDS + UGDS_WHITE + UGDS_BLACK + UGDS_HISP + UGDS_ASIAN + UGDS_AIAN + UGDS_NHPI + UGDS_2MOR + UGDS_NRA + UGDS_UNKN + COSTT4_A + AVGFACSAL + PCTPELL + AGE_ENTRY + FAMINC + ENDOWBEGIN + SAT_AVG + rank, data = data_set)

tidy(data_set_fit)
```

1.  We created a model based on the numerical variables that we chose for the data and created an additive model that considered how the variables influenced the rank of the college. At a first glance, variables with negative slopes mean that they have a negative correlation to ranking. For example, the higher the \`SAT_AVG\`, the lower the college is ranked on a numerical scale. So the college is ranked higher.

2.  Remove variables with the lowest slope coefficients until we get the best adjusted r-squared for the overall models.

```{r}
data_set_fit_true <- linear_reg() |>
  fit(rank ~ ADM_RATE + UGDS + UGDS_WHITE + UGDS_BLACK + UGDS_HISP + UGDS_ASIAN + UGDS_AIAN + UGDS_NHPI + UGDS_2MOR + UGDS_NRA + UGDS_UNKN + AVGFACSAL + PCTPELL + SAT_AVG + rank, data = data_set)

tidy(data_set_fit_true)
glance(data_set_fit_true)
```

We removed variables that had a slope with an absolute value less than 10 because that would mean that they have less of an impact on ranking than other variables.

1.  Look at slope coefficients for variables in models for variables in the model. The one with the highest slope is the most significant.

```{r}
data_set |>
  ggplot(aes(x = UGDS_WHITE, y = rank, color = REGION)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_x_continuous(limits = c(-2,2)) +
  labs(x = "Proportion of White-Identifying Undergraduate Students" ,
       y = "School Ranking")
```

## Results

Showcase how you arrived at answers to your research question using the techniques we have learned in class (and beyond, if you're feeling adventurous).

Our preliminary result shows that the number of undergraduate students that identify as white has major impact on the ranking of a college. Because of the positive slope, the more white students a college has, the number of the ranking is bigger which means a "lower" ranking.

Provide only the main results from your analysis. The goal is not to do an exhaustive data analysis (calculate every possible statistic and perform every possible procedure for all variables). Rather, you should demonstrate that you are proficient at asking meaningful questions and answering them using data, that you are skilled in interpreting and presenting results, and that you can accomplish these tasks using R. More is not better.
