---
title: "College Ranking Predictions"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: false
message: false
warning: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
library(corrplot)
library(tidymodels)
```

## Introduction

### Research Background

Every year, millions of students apply to colleges across the United States, and many of them use college rankings lists from sources such as *US News and World Report*, Forbes.com, and Niche.com to to help them decide where to apply and where to go. In recent years, these lists have been heavily criticized for focusing on "exclusivity and resources, rather than accessibility and economic mobility" (1).The system can also be easily manipulated by a university if that university prioritize certain metrics to raise their rankings, as seen in Northeastern University meteoric rise from #163 to #49 on the US News and World Report list in only 17 years. The president of Northeastern University even explicitly stated that it was a top priority of the university to raise its ranking (2).

\(1\) https://thehill.com/changing-america/enrichment/education/3641004-the-scandal-facing-college-ranking-lists-explained/

\(2\) https://www.bostonmagazine.com/news/2014/08/26/how-northeastern-gamed-the-college-rankings/

These lists are important because students applying to college trust these rankings and weigh them into their college decisions. Due to the large impact that a college has on a student's life, it is important to know where these rankings come from and what they actually measure. In this project, we will explore how influential different metrics are in determining a college's ranking.

In general, we want to examine how different variables affect a school's ranking on the **Niche College Ranking List**. We plan to look at variables that are typically thought to influence school rank such as average SAT score and admission rate, but we also want to look at variables that aren't typically thought of such as geographic region or endowment size.

### Research Questions and Hypothesis

**Question:** What variables (and what types of variables) have the greatest influence on a school's place in the Niche College Ranking list?

**Hypothesis:** We hypothesize that: ACT/SAT test score, acceptance rate, and median earnings 10 years after graduation will have the biggest impact on college rank.

In our project, we will join two data sets: The Niche College Rankings list and the US Department of Education College Scorecard

## Data

#### Data Set #1: Niche

-   The first data set comes from Niche's "2023 Best Colleges in America" list

    (https://www.niche.com/colleges/search/best-colleges/)

-   Niche aggregates data from a variety of sources, including the US Department of Education and reviews from students and alumni, to build their list of college rankings. The rankings list is updated monthly to reflect new data that Niche receives. However, Niche only receives data from the US Department of Education on an annual basis. The Niche data was scraped by Maia on October 17-19 2022.

-   There are 500 observations, representing the top 500 schools in the United States. Each observation has two variables: `college` and `rank`.

#### Data Set #2: US Department of Education

-   The second data set comes from the US Department of Education's College Scorecard, which is an exhaustive summary of characteristics and statistics for all colleges and universities in the United States.

    (https://collegescorecard.ed.gov/data/)

-   The College Scorecard is updated by the Education Department as it collects new data. Data used in the scorecard comes comes from data reported by the institutions, data on federal financial aid, data from taxes, and data from other federal agencies.

-   There were 2,989 variables in the original data set, many of which we don't need to answer our question, and since this data set was too large to load into RStudio, we used Excel to narrow this to 63 variables. There are 6681 observations in the data set, representing all of the colleges and universities in the United States.

### Data summary

Here is a summary of the variables we will be using in our analysis.

+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| Variable Name | Categorical (C)  | Description                                                                                                        | Levels of Cat. Variable                                                                                     |
|               |                  |                                                                                                                    |                                                                                                             |
|               | Quantitative (Q) |                                                                                                                    |                                                                                                             |
+===============+==================+====================================================================================================================+=============================================================================================================+
| college       | C                | Institution name                                                                                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| rank          | Q                | Niche rank                                                                                                         |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| REGION        | C                |                                                                                                                    | New England, Mid East, Great Lakes, Plains, Southeast, Southwest, Rocky Mountains, Far West, Outlying Areas |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACCREDAGENCY  | C                | Accreditor for Institution                                                                                         |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| CONTROL       | C                |                                                                                                                    | Public, Private nonprofit, private for-profit                                                               |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| CCBASIC       | C                | Carnegie Classification \--basic                                                                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ADM_RATE      | Q                | Admission rate                                                                                                     |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS          | Q                | Enrollment of undergraduate certificate/degree-seeking students                                                    |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_WHITE    | Q                | Total share of enrollment of undergraduate degree-seeking students who are white                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_BLACK    | Q                | Total share of enrollment of undergraduate degree-seeking students who are black                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_HISP     | Q                | Total share of enrollment of undergraduate degree-seeking students who are Hispanic                                |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_ASIAN    | Q                | Total share of enrollment of undergraduate degree-seeking students who are Asian                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_AIAN     | Q                | Total share of enrollment of undergraduate degree-seeking students who are American Indian/Alaska Native           |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_NHPI     | Q                | Total share of enrollment of undergraduate degree-seeking students who are Native Hawaiian/Pacific Islander        |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_2MOR     | Q                | Total share of enrollment of undergraduate degree-seeking students who are two or more races                       |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_NRA      | Q                | Total share of enrollment of undergraduate degree-seeking students who are non-resident aliens                     |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| UGDS_UNKN     | Q                | Total share of enrollment of undergraduate degree-seeking students whose race is unknown                           |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| NPT4_PUB      | Q                | Average net price for Title IV institutions (public institutions)                                                  |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| NPT4_PRIV     | Q                | Average net price for Title IV institutions (private for-profit and nonprofit institutions)                        |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| COSTT4_A      | Q                | Average cost of attendance (academic year institutions)                                                            |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| COSTT4_P      | Q                | Average cost of attendance (program-year institutions)                                                             |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| AVGFACSAL     | Q                | Average faculty salary                                                                                             |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| PCTPELL       | Q                | Percentage of undergraduates who receive a Pell Grant                                                              |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| C150_4        | Q                | Completion rate for first-time, full-time students at four-year institutions (150% of expected time to completion) |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| AGE_ENTRY     | Q                | Average age of entry                                                                                               |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| FEMALE        | Q                | Share of female students                                                                                           |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| MARRIED       | Q                | Share of married students                                                                                          |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| FIRST_GEN     | Q                | Share of first-generation students                                                                                 |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| FAMINC        | Q                | Average family income                                                                                              |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| MD_FAMINC     | Q                | Median family income                                                                                               |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ENDOWBEGIN    | Q                | Value of school's endowment at the beginning of the fiscal year                                                    |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| SAT_AVG       | Q                | Average SAT equivalent score of students admitted                                                                  |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| SATVRMID      | Q                | Midpoint of SAT scores at the institution (critical reading)                                                       |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| SATMTMID      | Q                | Midpoint of SAT scores at the institution (math)                                                                   |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| SATWRMID      | Q                | Midpoint of SAT scores at the institution (writing)                                                                |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACTCMMID      | Q                | Midpoint of the ACT cumulative score                                                                               |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACTENMID      | Q                | Midpoint of the ACT english score                                                                                  |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACTMTMID      | Q                | Midpoint of the ACT math score                                                                                     |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+
| ACTWRMID      | Q                | Midpoint of the ACT writing score                                                                                  |                                                                                                             |
+---------------+------------------+--------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------+

## Methodology

### Data Preparation

1.  To get the data, we scraped from Niche.com and downloaded data from the US Department of Education, and imported. The steps were done in an R script titled `niche-scrape.R`

```{r}
#| label: import
niche_data_500 <- read_csv("data/niche_data_500.csv")
us_dep_of_ed <- read_csv("data/us_dep_of_ed.csv", na = c("NULL", "PrivacySuppressed"))
```

2.  Some of the college names were slightly different between datasets, so we had to individually change names in the Department of Education dataset to match those in the Niche one before joining the two. The Observations for `University of South Florida - Sarasota-Manatee` and `University of South Florida - St. Petersburg` have been dropped from the `colleges` data set due to their non-existence in the `us_dep_of_ed` data set. They were not present in the data set because these two universities were combined with the main University of South Florida campus. This is why the `colleges` data set only has 498 observations.

```{r}
#| label: fix-college-names

us_dep_of_ed <- us_dep_of_ed |>
  mutate(
    INSTNM = case_when(
      INSTNM == "Columbia University in the City of New York" ~
        "Columbia University",

      INSTNM == "Washington University in St Louis" ~
        "Washington University in St. Louis",

      INSTNM == "University of California-Los Angeles" ~
        "University of California - Los Angeles",

      INSTNM == "University of Michigan-Ann Arbor" ~
        "University of Michigan - Ann Arbor",

      INSTNM == "Georgia Institute of Technology-Main Campus" ~
        "Georgia Institute of Technology",

      INSTNM == "University of Virginia-Main Campus" ~
        "University of Virginia",

      INSTNM == "United States Military Academy" ~
        "United States Military Academy at West Point",

      INSTNM == "The University of Texas at Austin" ~
        "University of Texas - Austin",

      INSTNM == "University of California-Berkeley" ~
        "University of California - Berkeley",

      INSTNM == "University of California-Irvine" ~
        "University of California - Irvine",

      INSTNM == "University of California-San Diego" ~
        "University of California - San Diego",

      INSTNM == "University of California-Davis" ~
        "University of California - Davis",

      INSTNM == "University of California-Santa Barbara" ~
        "University of California - Santa Barbara",

      INSTNM == "University of California-Santa Cruz" ~
        "University of California - Santa Cruz",

      INSTNM == "University of California-Riverside" ~
        "University of California - Riverside",

      INSTNM == "University of North Carolina Wilmington" ~
        "University of North Carolina - Wilmington",

      INSTNM == "University of North Carolina at Greensboro" ~
        "University of North Carolina - Greensboro",

      INSTNM == "Albany College of Pharmacy and Health Sciences" ~
        "Albany College of Pharmacy & Health Sciences",

      INSTNM == "Arizona State University Campus Immersion" ~
        "Arizona State University",

      INSTNM == "Arizona State University-Downtown Phoenix" ~
        "Arizona State University - Downtown Phoenix Campus",

      INSTNM == "Augustana College" ~
        "Augustana College - Illinois",

      UNITID == "173160" ~
        "Bethel University - Minnesota",

      INSTNM == "Binghamton University" ~
        "Binghamton University, SUNY",

      INSTNM == "Bowling Green State University-Main Campus" ~
        "Bowling Green State University",

      INSTNM == "California Polytechnic State University-San Luis Obispo" ~
        "California Polytechnic State University (Cal Poly) - San Luis Obispo",

      INSTNM == "California State University-Fullerton" ~
        "California State University - Fullerton",

      INSTNM == "California State University-Long Beach" ~
        "California State University - Long Beach",

      INSTNM == "The College of Wooster" ~
        "College of Wooster",

      INSTNM == "Colorado State University-Fort Collins" ~
        "Colorado State University",

      INSTNM == "Concordia University-Wisconsin" ~
        "Concordia University - Wisconsin",

      INSTNM == "CUNY Bernard M Baruch College" ~
        "CUNY Baruch College",

      INSTNM == "CUNY City College" ~
        "CUNY City College of New York",

      INSTNM == "D'Youville College" ~
        "D'Youville",

      INSTNM == "Eastern New Mexico University-Main Campus" ~
        "Eastern New Mexico University",

      INSTNM == "Embry-Riddle Aeronautical University-Daytona Beach" ~
        "Embry-Riddle Aeronautical University - Daytona Beach",

      INSTNM == "Embry-Riddle Aeronautical University-Worldwide" ~
        "Embry-Riddle Aeronautical University - Worldwide",

      INSTNM == "Florida Agricultural and Mechanical University" ~
        "Florida A&M University",

      INSTNM == "Franklin and Marshall College" ~
        "Franklin & Marshall College",

      INSTNM == "Hobart William Smith Colleges" ~
        "Hobart and William Smith Colleges",

      INSTNM == "Indiana University-Bloomington" ~
        "Indiana University - Bloomington",

      INSTNM == "Indiana University-East" ~
        "Indiana University - East",

      INSTNM == "Indiana University-Purdue University-Indianapolis" ~
        "Indiana University-Purdue University - Indianapolis (IUPUI)",

      INSTNM == "Indiana Wesleyan University-Marion" ~
        "Indiana Wesleyan University",

      INSTNM == "Keiser University-Ft Lauderdale" ~
        "Keiser University - Fort Lauderdale",

      INSTNM == "Kent State University at Kent" ~
        "Kent State University",

      INSTNM == "Louisiana State University and Agricultural & Mechanical College" ~
        "Louisiana State University",

      UNITID == "151786" ~
        "Marian University Indianapolis",

      INSTNM == "Maryville University of Saint Louis" ~
        "Maryville University",

      INSTNM == "Metropolitan State University" ~
        "Metropolitan State University - Minnesota",

      INSTNM == "Miami University-Oxford" ~
        "Miami University",

      INSTNM == "Minnesota State University-Mankato" ~
        "Minnesota State University, Mankato",

      INSTNM == "Molloy College" ~
        "Molloy University",

      INSTNM == "Monroe College" ~
        "Monroe College - Bronx/New Rochelle",

      INSTNM == "Mount Saint Mary's University" ~
        "Mount Saint Mary's University Los Angeles",

      INSTNM == "New Mexico State University-Main Campus" ~
        "New Mexico State University",

      INSTNM == "New Mexico Institute of Mining and Technology" ~
        "New Mexico Tech",

      INSTNM == "North Carolina State University at Raleigh" ~
        "North Carolina State University",

      INSTNM == "North Dakota State University-Main Campus" ~
        "North Dakota State University",

      UNITID ==  "154101" ~
        "Northwestern College - Iowa",

      INSTNM == "Ohio University-Main Campus" ~
        "Ohio University",

      INSTNM == "Oklahoma State University-Main Campus" ~
        "Oklahoma State University",

      INSTNM == "Pacific University" ~
        "Pacific University Oregon",

      INSTNM == "The Pennsylvania State University" ~
        "Penn State",

      INSTNM == "Pennsylvania State University-Penn State Fayette- Eberly" ~
        "Penn State Fayette, The Eberly Campus",

      INSTNM == "Pennsylvania State University-Penn State York" ~
        "Penn State York",

      INSTNM == "Purdue University-Main Campus" ~
        "Purdue University",

      INSTNM == "Rutgers University-New Brunswick" ~
        "Rutgers University - New Brunswick",

      INSTNM == "Rutgers University-Newark" ~
        "Rutgers University - Newark",

      INSTNM == "The University of the South" ~
        "Sewanee - The University of the South",

      INSTNM == "South Dakota School of Mines and Technology" ~
        "South Dakota School of Mines & Technology",

      INSTNM == "St Bonaventure University" ~
        "St. Bonaventure University",

      INSTNM == "Saint John Fisher College" ~
        "St. John Fisher University",

      INSTNM == "St. Joseph's University-New York" ~
        "St. Joseph's University - New York",

      INSTNM == "St Lawrence University" ~
        "St. Lawrence University",

      INSTNM == "St Olaf College" ~
        "St. Olaf College",

      INSTNM == "Stanbridge University" ~
        "Stanbridge University - Orange County",

      INSTNM == "Stephen F Austin State University" ~
        "Stephen F. Austin State University",

      INSTNM == "Stony Brook University" ~
        "Stony Brook University, SUNY",

      INSTNM == "SUNY College of Environmental Science and Forestry" ~
        "SUNY College of Environmental Science & Forestry",

      INSTNM == "Farmingdale State College" ~
        "SUNY Farmingdale State College",

      INSTNM == "State University of New York at New Paltz" ~
        "SUNY New Paltz",

      INSTNM == "Texas A & M University-College Station" ~
        "Texas A&M University",

      INSTNM == "American Musical and Dramatic Academy" ~
        "The American Musical and Dramatic Academy (AMDA) - New York",

      INSTNM == "The College of Saint Scholastica" ~
        "The College of St. Scholastica",

      INSTNM == "Cooper Union for the Advancement of Science and Art" ~
        "The Cooper Union for the Advancement of Science and Art",

      INSTNM == "The Master's University and Seminary" ~
        "The Master's University",

      INSTNM == "Ohio State University-Main Campus" ~
        "The Ohio State University",

      INSTNM == "University of Alabama in Huntsville" ~
        "The University of Alabama in Huntsville",

      INSTNM == "University of Baltimore" ~
        "The University of Baltimore",

      INSTNM == "University of Tulsa" ~
        "The University of Tulsa",

      INSTNM == "University of Virginia's College at Wise" ~
        "The University of Virginia's College at Wise",

      INSTNM == "Trinity College" ~
        "Trinity College - Connecticut",

      INSTNM == "Tulane University of Louisiana" ~
        "Tulane University",

      UNITID == "196866" ~
        "Union College - New York",

      INSTNM == "University at Buffalo" ~
        "University at Buffalo, SUNY",

      INSTNM == "University of Alabama at Birmingham" ~
        "University of Alabama - Birmingham",

      INSTNM == "University of Cincinnati-Main Campus" ~
        "University of Cincinnati",

      INSTNM == "University of Colorado Denver/Anschutz Medical Campus" ~
        "University of Colorado Denver",

      INSTNM == "University of Maryland-College Park" ~
        "University of Maryland - College Park",

      INSTNM == "University of Maryland-Baltimore County" ~
        "University of Maryland, Baltimore County",

      INSTNM == "University of Massachusetts-Amherst" ~
        "University of Massachusetts - Amherst",

      INSTNM == "University of Massachusetts-Lowell" ~
        "University of Massachusetts Lowell",

      INSTNM == "University of Michigan-Dearborn" ~
        "University of Michigan - Dearborn",

      INSTNM == "University of Minnesota-Crookston" ~
        "University of Minnesota Crookston",

      INSTNM == "University of Minnesota-Duluth" ~
        "University of Minnesota Duluth",

      INSTNM == "University of Minnesota-Twin Cities" ~
        "University of Minnesota Twin Cities",

      INSTNM == "University of Missouri-Columbia" ~
        "University of Missouri",

      INSTNM == "Embry-Riddle Aeronautical University-Prescott" ~
        "Embry-Riddle Aeronautical University - Prescott",

      INSTNM == "South Dakota State University" ~
        "South Dakota State University",

      INSTNM == "St Catherine University" ~
        "St. Catherine University",

      INSTNM == "University of Missouri-Kansas City" ~
        "University of Missouri - Kansas City",

      INSTNM == "University of Missouri-St Louis" ~
        "University of Missouri - St. Louis",

      INSTNM == "University of Nebraska-Lincoln" ~
        "University of Nebraska - Lincoln",

      INSTNM == "University of Nevada-Reno" ~
        "University of Nevada - Reno",

      INSTNM == "University of New Hampshire-Main Campus" ~
        "University of New Hampshire",

      INSTNM == "University of Oklahoma-Norman Campus" ~
        "University of Oklahoma",

      INSTNM == "University of Pittsburgh-Pittsburgh Campus" ~
        "University of Pittsburgh",

      INSTNM == "University of South Carolina-Columbia" ~
        "University of South Carolina",

      INSTNM == "University of St Francis" ~
        "University of St. Francis - Illinois",

      UNITID == "174914" ~
        "University of St. Thomas - Minnesota",

      UNITID == "227863" ~
        "University of St. Thomas - Texas",

      INSTNM == "The University of Tennessee-Knoxville" ~
        "University of Tennessee",

      INSTNM == "The University of Tennessee-Martin" ~
        "University of Tennessee at Martin",

      INSTNM == "The University of Texas at Arlington" ~
        "University of Texas - Arlington",

      INSTNM == "The University of Texas at Dallas" ~
        "University of Texas - Dallas",

      INSTNM == "The University of Texas Permian Basin" ~
        "University of Texas - Permian Basin",

      INSTNM == "The University of Texas Rio Grande Valley" ~
        "University of Texas - Rio Grande Valley",

      INSTNM == "The University of Texas at Tyler" ~
        "University of Texas - Tyler",

      INSTNM == "University of Washington-Seattle Campus" ~
        "University of Washington",

      INSTNM == "University of Washington-Bothell Campus" ~
        "University of Washington - Bothell",

      INSTNM == "University of Washington-Tacoma Campus" ~
        "University of Washington - Tacoma",

      INSTNM == "The University of West Florida" ~
        "University of West Florida",

      INSTNM == "University of Wisconsin-Madison" ~
        "University of Wisconsin",

      INSTNM == "University of Wisconsin-La Crosse" ~
        "University of Wisconsin - La Crosse",

      INSTNM == "Virginia Polytechnic Institute and State University" ~
        "Virginia Tech",

      INSTNM == "West Texas A & M University" ~
        "West Texas A&M University",

      UNITID == "230807" ~
        "Westminster College - Utah",

      INSTNM == "Wheaton College" ~
        "Wheaton College - Illinois",

      INSTNM == "Wheaton College (Massachusetts)" ~
        "Wheaton College - Massachusetts",

      TRUE ~ INSTNM
    )
  )
```

3.  We joined the US Department of Education dataset to the Niche dataset by the college names. The first 10 rows are displayed below.

```{r}
#| label: join-datasets

colleges <- niche_data_500 |>
  left_join(us_dep_of_ed, by = c("college" = "INSTNM")) |>
  filter(rank != 141) |>
  filter(rank != 181)

colleges |>
  slice(1:10) |>
  knitr::kable()
```

4.  Then, we selected the variables that we thought could have an impact on college ranking as a starting point for our analysis. These are listed in our data summary.

```{r}
#| label: select-variables

chosen_variables <- c(
  "college",
  "rank",
  "REGION",
  "ACCREDAGENCY",
  "CONTROL",
  "CCBASIC",
  "ADM_RATE",
  "UGDS",
  "UGDS_WHITE",
  "UGDS_BLACK",
  "UGDS_HISP",
  "UGDS_ASIAN",
  "UGDS_AIAN",
  "UGDS_NHPI",
  "UGDS_2MOR",
  "UGDS_NRA",
  "UGDS_UNKN",
  "NPT4_PUB",
  "NPT4_PRIV",
  "COSTT4_A",
  "COSTT4_P",
  "AVGFACSAL",
  "PCTPELL",
  "C150_4",
  "AGE_ENTRY",
  "FEMALE",
  "MARRIED",
  "FIRST_GEN",
  "FAMINC",
  "MD_FAMINC",
  "ENDOWBEGIN",
  "SAT_AVG",
  "SATVRMID",
  "SATMTMID",
  "SATWRMID",
  "ACTCMMID",
  "ACTENMID",
  "ACTMTMID",
  "ACTWRMID"
)

data <- colleges |>
  select(chosen_variables)
```

5.  Some of the categorical variables in the US Department of Education dataset (`REGION`, `CONTROL`, and `CCBASIC`) used numbers to represent the different levels, so we looked at the data dictionary and replaced each number with the words that it represents.

```{r}
#| label: clean-categorical-variables

data <- data |>
  mutate(
    REGION = as.character(REGION),
    REGION = case_when(
      REGION == 1 ~ "New England",
      REGION == 2 ~ "Mid East",
      REGION == 3 ~ "Great Lakes",
      REGION == 4 ~ "Plains",
      REGION == 5 ~ "Southeast",
      REGION == 6 ~ "Southwest",
      REGION == 7 ~ "Rocky Mountains",
      REGION == 8 ~ "Far West",
      REGION == 9 ~ "Outlying Areas"
    ),
    CONTROL = as.character(CONTROL),
    CONTROL = case_when(
      CONTROL == 1 ~ "Public",
      CONTROL == 2 ~ "Private, Nonprofit",
      CONTROL == 3 ~ "Proprietary"
    ),
    CCBASIC = as.character(CCBASIC),
    CCBASIC = case_when(
      CCBASIC == -2 ~ "Not applicable",
      CCBASIC == 0 ~ "Not classified",
      CCBASIC == 1 ~ "Associate's Colleges: High Transfer-High Traditional",
      CCBASIC == 2 ~ "Associate's Colleges: High Transfer-Mixed Traditional/Nontraditional",
      CCBASIC == 3 ~ "Associate's Colleges: High Transfer-High Nontraditional",
      CCBASIC == 4 ~ "Associate's Colleges: Mixed Transfer/Career & Technical-High Traditional",
      CCBASIC == 5 ~ "Associate's Colleges: Mixed Transfer/Career & Technical-Mixed Traditional/Nontraditional",
      CCBASIC == 6 ~ "Associate's Colleges: Mixed Transfer/Career & Technical-High Nontraditional",
      CCBASIC == 7 ~ "Associate's Colleges: High Career & Technical-High Traditional",
      CCBASIC == 8 ~ "Associate's Colleges: High Career & Technical-Mixed Traditional/Nontraditional",
      CCBASIC == 9 ~ "Associate's Colleges: High Career & Technical-High Nontraditional",
      CCBASIC == 10 ~ "Special Focus Two-Year: Health Professions",
      CCBASIC == 11 ~ "Special Focus Two-Year: Technical Professions",
      CCBASIC == 12 ~ "Special Focus Two-Year: Arts & Design",
      CCBASIC == 13 ~ "Special Focus Two-Year: Other Fields",
      CCBASIC == 14 ~ "Baccalaureate/Associate's Colleges: Associate's Dominant",
      CCBASIC == 15 ~ "Doctoral Universities: Very High Research Activity",
      CCBASIC == 16 ~ "Doctoral Universities: High Research Activity",
      CCBASIC == 17 ~ "Doctoral/Professional Universities",
      CCBASIC == 18 ~ "Master's Colleges & Universities: Larger Programs",
      CCBASIC == 19 ~ "Master's Colleges & Universities: Medium Programs",
      CCBASIC == 20 ~ "Master's Colleges & Universities: Small Programs",
      CCBASIC == 21 ~ "Baccalaureate Colleges: Arts & Sciences Focus",
      CCBASIC == 22 ~ "Baccalaureate Colleges: Diverse Fields",
      CCBASIC == 23 ~ "Baccalaureate/Associate's Colleges: Mixed Baccalaureate/Associate's",
      CCBASIC == 24 ~ "Special Focus Four-Year: Faith-Related Institutions",
      CCBASIC == 25 ~ "Special Focus Four-Year: Medical Schools & Centers",
      CCBASIC == 26 ~ "Special Focus Four-Year: Other Health Professions Schools",
      CCBASIC == 27 ~ "Special Focus Four-Year: Engineering Schools",
      CCBASIC == 28 ~ "Special Focus Four-Year: Other Technology-Related Schools",
      CCBASIC == 29 ~ "Special Focus Four-Year: Business & Management Schools",
      CCBASIC == 30 ~ "Special Focus Four-Year: Arts, Music & Design Schools",
      CCBASIC == 31 ~ "Special Focus Four-Year: Law Schools",
      CCBASIC == 32 ~ "Special Focus Four-Year: Other Special Focus Institutions",
      CCBASIC == 33 ~ "Tribal Colleges"
    )
  )
```

6.  All of the numerical variables are on different scales---for example, SAT scores can range from 400 to 1600, while admittance rate can only range from 0 to 1---and so we standardized them for easier analysis and comparison. We used the `scale()` functions (found at https://www.statology.org/standardize-data-in-r/) to make the mean value of each numeric variable 0, and the standard deviation 1. The first 10 rows of our standardized dataset are printed below.

```{r}
#| label: scale-numerical-variables

scaled_continuous_numeric_variables <- data |>
  select_if(is.numeric) |>
  select(-rank) |>
  scale() |>
  data.frame()

other_variables <- data |>
  select(rank | REGION | CONTROL | CCBASIC | !where(is.numeric))

scaled_data <- cbind(other_variables, scaled_continuous_numeric_variables) |>
  relocate(college)

scaled_data |>
  slice(1:10) |>
  knitr::kable()
```

### Preliminary Exploration and Visualization

1.  We broke down the data into 5 different groups based on the college rank, and then took the mean of selected variables for each of the groups. This gave us an idea of how different variables change across different types of school.

```{r}
#| label: exploration
colleges_levels <- colleges |> 
  mutate(
    level = case_when(
      rank <= 100 ~ "1 to 100",
      rank > 100 & rank <= 200 ~ "101 to 200",
      rank > 200 & rank <= 300 ~ "201 to 300",
      rank > 300 & rank <= 400 ~ "301 to 400",
      rank > 400 & rank <= 498 ~ "401 to 498",
    )
  )

colleges_levels |> 
  group_by(level) |> 
  summarize(
    mean_ADM_Rate = mean(ADM_RATE, na.rm = TRUE),
    mean_SAT_AVG = mean(SAT_AVG, na.rm = TRUE),
    mean_ACTCMMID = mean(ACTCMMID, na.rm = TRUE),
    mean_UGDS_WHITE = mean(UGDS_WHITE, na.rm = TRUE),
    mean_UGDS_ASIAN = mean(UGDS_ASIAN, na.rm = TRUE),
    mean_COSTT4_A = mean(COSTT4_A, na.rm = TRUE)
  ) |>
  knitr::kable()
```

2.  We have plotted rank versus several of the numerical variables. This gives us an idea of how different metrics affect college rank. Just by looking at the graphs, Admission Rate, Mean SAT, Median ACT, Average Faculty Salary, and Graduation Rate seem to have the strongest correlation.

```{r}
#| label: graphs

scaled_data|> 
  ggplot(aes(x = ADM_RATE, y = rank, color = CONTROL)) +
  geom_point() +
  labs(
    x = "Admission Rate",
    y = "Rank",
    color = "Public/Private", 
    title = "Rank vs Admission Rate by Public/Private"
  )

colleges |> 
  ggplot(aes(x = UGDS_WHITE, y = rank)) +
  geom_point() +
  labs(
    x = "% of Undergrads Who are White",
    y = "Rank",
    title = "Rank vs % of Undergrads Who are White"
  )
  

colleges |> 
  ggplot(aes(x = UGDS_BLACK, y = rank)) +
  geom_point() +
  labs(
    x = "% of Undergrads Who are Black",
    y = "Rank",
    title = "Rank vs % of Undergrads Who are Black"
  )

scaled_data |> 
  ggplot(aes(x = COSTT4_A, y = rank, color = CONTROL)) +
  geom_point() +
  labs(
    x = "Cost of Attendance", 
    y = "Rank",
    color = "Public/Private",
    title = "Rank vs Cost of Attendance by Public/Private"
  ) 

scaled_data |> 
  ggplot(aes(x = AVGFACSAL, y = rank)) +
  geom_point() +
  labs(
    x = "Average Faculty Salary", 
    y = "Rank",
    title = "Rank vs Average Faculty Salary"
  ) 

scaled_data |> 
  ggplot(aes(x = C150_4, y = rank)) +
  geom_point() +
  labs(
    x = "Graduation Rate", 
    y = "Rank",
    title = "Rank vs Graduation Rate"
  ) 

scaled_data |> 
  ggplot(aes(x = ENDOWBEGIN, y = rank)) +
  geom_point() +
  labs(
    x = "Endowment Size", 
    y = "Rank",
    title = "Rank vs Endowment Size"
  ) 

colleges |> 
  ggplot(aes(x = SAT_AVG, y = rank)) +
  geom_point() +
  labs(
    x = "Mean SAT", 
    y = "Rank",
    title = "Rank vs Mean SAT"
  ) 

colleges |> 
  ggplot(aes(x = ACTCMMID, y = rank)) +
  geom_point() +
  labs(
    x = "Median ACT", 
    y = "Rank",
    title = "Rank vs Median ACT"
  ) 
```

3.  We checked the correlation coefficients between the variables so we don't use variables that are too similar in our model. For the ones that have a absolute value of r greater than 0.8, we picked only one to put in the model. We removed the columns that has more than 100 NA and used the rest for calculating correlations.

    ```{r}
    #| label: correlation-matrix

    data_for_corr <- scaled_data |>
      select(is.numeric)

    correlations = data_for_corr |>
      select(-c("NPT4_PUB", "NPT4_PRIV", "COSTT4_P", "SATWRMID", "SATVRMID", "SATMTMID", "ACTENMID", "ACTMTMID", "ACTWRMID")) |>
        na.omit() |>
        cor() 

    correlations |>
        corrplot(method="color")

    correlations
    ```
According to the correlation matrix, we need to remove variable pairs that have a high correlation (the absolute value \> 0.8). In order to do so, we filter the following correlation table to only display entries with value greater or equal to 0.8.

```{r} 
#| label: display-correlation-pair-to-be-removed

correlations[correlations < 0.8 | correlations ==1] <- ""

correlations |> knitr::kable() 
```


Accoridng to the filtered table, these variable pairs are (MD_FAMINC, FAMINC), (C150_4, SAT_AVG), (C150_4, ACTCMMID) and (ACTCMMID, SAT_AVG). Therefore, we will drop the variables C150_4, MD_FAMINC, ACTCMMID and preserve SAT_AVG and FAMINC to represent all other variables.

### Build model

We began our modeling by putting all of our chosen variables into a model

```{r}
#| label: glimpsed-data

data_set <- scaled_data |>
  select(c("ADM_RATE", "UGDS", "UGDS_WHITE", "UGDS_BLACK", "UGDS_HISP", "UGDS_ASIAN", "UGDS_AIAN", "UGDS_NHPI", "UGDS_2MOR", "UGDS_NRA", "UGDS_UNKN", "COSTT4_A", "AVGFACSAL", "PCTPELL", "AGE_ENTRY", "FAMINC", "ENDOWBEGIN", "SAT_AVG", "rank", "REGION", "CONTROL", "CCBASIC", "college", "ACCREDAGENCY", "FEMALE", "MARRIED", "FIRST_GEN"))

glimpse(data_set)
```

Then we need to split dataset into two parts: a training and a test set. We will use 80% of the college data for trainning our model and the rest 20% for testing our model.

```{r}
#| label: split-data

set.seed(777)

# Build training data set
train <- data_set |>
  sample_frac(.80)

# Build testing data set
test <- anti_join(data_set, train, by = 'rank')
```


#### Iteration #0

We begin with creating the first model based on the numerical variables that we chose for the data and created an additive model that considered how the variables influenced the rank of the college. 

```{r}
#| label: linear-reg-data

fit1 <- linear_reg() |>
  fit(rank ~ ADM_RATE + UGDS + UGDS_WHITE + UGDS_BLACK + UGDS_HISP + UGDS_ASIAN + UGDS_AIAN + UGDS_NHPI + UGDS_2MOR + UGDS_NRA + UGDS_UNKN + COSTT4_A + AVGFACSAL + PCTPELL + AGE_ENTRY + FAMINC + ENDOWBEGIN + SAT_AVG , data = train)

tidy(fit1) |> knitr::kable() 
```


At a first glance, variables with negative slopes mean that they have a negative correlation to ranking. For example, the higher the \`SAT_AVG\`, the lower the college is ranked on a numerical scale. So the college is technically ranked higher. 

Variables UGDS_NRA, UGDS_UNKN, COSTT4_A, PCTPELL, AGE_ENTRY, UGDS_WHITE, UGDS_BLACK, UGDS_HISP, UGDS_ASIAN, UGDS_AIAN, UGDS_NHPI, UGDS_2MOR and ENDOWBEGIN are have a relative high p-value around 0.3, which is much greater than 0.05. Therefore, we need to drop these variables for the next model iteration.

Then we can evaluation our first primitive model by Rsqure:

```{r}
#| label: evaluation1

test_results <- predict(fit1, new_data = test) |>
                            bind_cols(test)


# R2 on test set
rsq(test_results,
    truth = test$rank,
    estimate = .pred) |> knitr::kable() 
```
As we can see the model R square is around 70%. 

#### Iteration #1

First we need to avoid using the variables with high p-values as mentioned above to train a new model.

```{r}
#| label: fit2

fit2 <- linear_reg() |>
  fit(rank ~ ADM_RATE + UGDS + AVGFACSAL + SAT_AVG, data = train)

tidy(fit2) |> knitr::kable() 
```
Then we can evaluation our second model by Rsqure:

```{r}
test_results <- predict(fit2, new_data = test) |>
                            bind_cols(test)

# R2 on test set
rsq(test_results,
    truth = test$rank,
    estimate = .pred) |> knitr::kable() 
```

Since the rsq value is 0.68 which is even smaller than 0.69 of our first model, we need to implement forward selection by adding variables back to our model.

#### Iteration #3

From our first and second iteration, AGE_ENTRY, among the variables removed, has the smallest p-value of 0.1. Therefore we are going to add AGE_ENTRY back to our model for the iteration.


```{r}
#| label: fit3

fit3 <- linear_reg() |>
  fit(rank ~ ADM_RATE + UGDS + AVGFACSAL + SAT_AVG + AGE_ENTRY, data = train)

tidy(fit3) |> knitr::kable() 
```

```{r}
test_results <- predict(fit3, new_data = test) |>
                            bind_cols(test)

# R2 on test set
rsq(test_results,
    truth = test$rank,
    estimate = .pred) |> knitr::kable() 
```

However the model accuracy is still not as good as the first model which means we need further forward selection by adding more variables back.

#### Iteration #4

From our first and second iteration, UGDS_UNKN, among the variables removed, has the smallest p-value of 0.310. Therefore we are going to add UGDS_UNKN back to our model for the iteration.

```{r}
#| label: fit4

fit4 <- linear_reg() |>
  fit(rank ~ ADM_RATE + UGDS + AVGFACSAL + SAT_AVG + AGE_ENTRY + UGDS_UNKN, data = train)

tidy(fit4) |> knitr::kable() 
```

```{r}
test_results <- predict(fit4, new_data = test) |>
                            bind_cols(test)

# R2 on test set
rsq(test_results,
    truth = test$rank,
    estimate = .pred) |> knitr::kable() 
```


## Results

Our preliminary result shows that the number of undergraduate students that identify as white has major impact on the ranking of a college. Because of the positive slope, the more white students a college has, the number of the ranking is bigger which corresponds to a "lower" ranking.
